\ProjectName{}: Data Imputation as a Query Optimization

### 2016-12-11 paper/algorithm.tex
paper: some cleanup, still missing motivating example and final experiments

Our optimizer searches a restricted space (Sec.~\ref{sec:search-space}) of query plans for a plan that minimizes a metric (Sec.~\ref{sec:cost-model}) which combines the runtime cost of the query and the quality of the results.
This plan must not emit any tuples which contain null values, regardless of the state of the base tables.
Additionally, the traditional relational operators (selection $\sigma$, projection $\pi$, join $\bowtie$, and group-by/aggregate) must never observe a null value.

@@

\subsection{Cost model}
\label{sec:cost-model}
We rank query plans $Q$ using a cost model $\text{Cost}(Q) = \alpha \times \text{Time}(Q) + (1 - \alpha) \times \text{Loss}(Q)$. $\text{Time}(Q)$ is an estimate of the runtime of the query which is derived from table statistics and selectivity estimation of the query predicates. $\text{Loss}(Q)$ is an estimate of the amount of error introduced by the imputation procedure. $\alpha$ is a parameter to the query optimizer that controls the emphasis on quality versus performance. $\alpha = 1.0$ means that the query should be as fast as possible, $\alpha=0.0$ means that the query should be as accurate as possible.

In order to correctly estimate $\text{Time}(Q)$ and $\text{Loss}(Q)$, the system must have cardinality estimates
for each sub-query in each query plan.
These cardinality estimates are impacted not just by filtering or joining, as in the traditional relational calculus, but also by the imputation operators.
For example, a drop operator will reduce the cardinality of the result while an impute operator will maintain the same cardinality as the input.
For simplicity, each of the logical nodes in a query plan refers to a set of histograms.
When the optimizer creates a new query plan, it copies the histograms of the sub-plans and modifies them as necessary to account for new operation in the plan.
Algorithm~\ref{algo:histogram-transformation} describes the process of generating new histograms from sub-plans.

@@

\todo{Micah: Explanation edited a bit. In the algorithm, when we impute attribute $d$, we
condition on attributes $C \cup D \backslash \{d\}$. Hopefully the way I note it is clear;
feel free to modify otherwise.}

>>>>

Our optimizer searches a restricted space (Sec.~\ref{sec:search-space}) of query plans for a plan that minimizes a metric (Sec.~\ref{sec:cost-model}) which combines the runtime cost of the query and the quality of the results.
This plan must not emit any tuples which contain null values, regardless of the state of the base tables.
Additionally, the traditional relational operators (selection $\sigma$, projection $\pi$, join $\bowtie$, and group-by/aggregate) must never observe a null value in any attribute that they directly operate on.

@@

\subsection{Cost model}
\label{sec:cost-model}
We rank query plans $Q$ using a cost model $\text{Cost}(Q) = (1 - \alpha) \times \text{Time}(Q) + \alpha \times \text{Loss}(Q)$. $\text{Time}(Q)$ is an estimate of the runtime of the query which is derived from table statistics and selectivity estimation of the query predicates. $\text{Loss}(Q)$ is an estimate of the amount of error introduced by the imputation procedure. $\alpha$ is a parameter to the query optimizer that controls the emphasis on quality over performance. $\alpha = 1.0$ means that the query should be as accurate as possible (at the expense of performance), $\alpha=0.0$ means that the query should be as fast
as possible (at the expense of accuracy).

In order to correctly estimate $\text{Time}(Q)$ and $\text{Loss}(Q)$, the system must have cardinality estimates
for each sub-query in each query plan.
These cardinality estimates are impacted not just by filtering or joining, as in the traditional relational calculus, but also by the imputation operators.
For example, a drop operator will reduce the cardinality of the result while an impute operator will maintain the same cardinality as the input.
For simplicity, each of the logical nodes in a query plan points to a set of histograms.
When the optimizer creates a new query plan, it copies the histograms of the sub-plans and modifies them as necessary to account for new operation in the plan.
Algorithm~\ref{algo:histogram-transformation} describes the process of generating new histograms from sub-plans.

@@

### 2016-12-11 paper/experiments.tex
paper: some cleanup, still missing motivating example and final experiments

Figure~\ref{fig:experiments} provides a summary of our experimental results. For purposes of simplicity, we present only the results associated with
queries that return a result size matching our base case. This addresses the fact that deriving a reasonable error metric for set-queries 
is inherently task specific (e.g. consider tuning precision vs. recall). Meanwhile, for all queries that match, we apply a simple sum of errors squared
(SSE). Figure~\ref{fig:experiments} indicates the query number and the error produced by a query engine with traditional SQL null semantics (base error).
For imputations, we present two extremes of the $\alpha$ parameter: 0 and 1, which reflect a a focus on runtime and information loss, respectively.
For each query, the table displays the new error ($\epsilon$), the planning time ($t_p$) and the running time ($t_r$) in seconds.

@@

Finally, we performed CERT imputation on the entirety of the ACS table, in order to compare our system to the traditional, pre-processing approach to imputation.
We note that the latter took \todo{add time}, compared to our maximum of \todo{add time}. Given the varying needs of database users, \ProjectName{} clearly allows flexibility without
a significant performance hit.

>>>>

Figure~\ref{fig:experiments} provides a summary of our experimental results. For purposes of simplicity, we present only the results associated with
queries that return a result size matching our base case. This addresses the fact that deriving a reasonable error metric for set-queries 
is inherently task specific (e.g. consider tuning precision vs. recall). Meanwhile, for all queries that match, we apply a simple sum of squared errors
(SSE). Figure~\ref{fig:experiments} indicates the query number and the error produced by a query engine with traditional SQL null semantics (base error).
For imputations, we present two extremes of the $\alpha$ parameter: 0 and 1, which reflect a a focus on runtime and information loss, respectively.
For each query, the table displays the new error ($\epsilon$), the planning time ($t_p$) and the running time ($t_r$) in seconds.

@@

Finally, we performed $CE-CART$ imputation on the entirety of the ACS table, in order to compare our system to the traditional, pre-processing approach to imputation.
We note that the latter took \todo{add time}, compared to our maximum of \todo{add time}. Given the varying needs of database users, \ProjectName{} clearly allows flexibility without
a significant performance hit.
